Incorporating periodic variability in hidden Markov models for animal movement

### 2016-01-28 paper3.Rnw
BMB beginning of updates

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{ HMM}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal heterogeneity and
autocorrelation (persistence) in group occupancy can obscure such
signals. We use finite mixture and hidden Markov models, two standard
clustering techniques, to model high-resolution hourly movement data
from Florida panthers. Allowing for temporal heterogeneity in
transition probabilities, a straightforward but rarely explored model
extension, resolves some shortcomings of current modeling frameworks
and clarifies the behavioural patterns of panthers.

@@

\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity}

@@

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
\todo{do we want to call these transition probabilities (throughout)? ML: DONE}
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in transitioning behaviour over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a very fast time scale, such
that movement behaviours at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.  Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transitioning
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transitioning probabilities -- in particular,
transitioning probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  We re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
analyze the movement of Florida panthers.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have decided \emph{a priori} on a
number of underlying states, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,
  fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,
kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not used
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

@@

\todo{what do we think of this subsection title? ML:sounds good to me}

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transitioning pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.

>>>>

\documentclass{article}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage[utf8]{inputenc} % for accented characters
%% stuff for editing
%\usepackage[markup=nocolor,addedmarkup=bf,deletedmarkup=sout]{changes}
%% to suppress notes & comments: \usepackage[final]{changes}
\usepackage[backgroundcolor=lightgray,textsize=tiny]{todonotes}
\usepackage{setspace}
\bibliographystyle{chicago}
\title{Incorporating periodic variability in hidden Markov models for animal movement}
\author{Michael Li and Benjamin M. Bolker }
\date{\today}

@@

Clustering time-series data into discrete groups can improve
prediction as well as providing insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
probability that individuals occupy or move between groups can obscure
such signals. We use finite mixture and hidden Markov models, two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers. Allowing for temporal
heterogeneity in transition probabilities, a straightforward but
rarely explored model extension, resolves some shortcomings of current
modeling frameworks and clarifies the behavioural patterns of
panthers. More generally, incorporating previously neglected
structure in statistical models can allow researchers to more
accurately identify models of appropriate complexity.

@@

\keywords{hidden Markov model, animal movement, temporal autocorrelation, temporal heterogeneity, Florida panther}

@@

HMMs have typically been used to model movements over short time
scales, where the probability of transitioning between movement states is
approximately constant. Changes in latent/hidden behavioural state/mode transition 
probabilities based on
the local environment can be accounted for incorporating environmental
covariates in the HMM \citep{patterson_classifying_2009}, or by more
\emph{ad hoc} comparisons between inferred states and environmental
conditions \citep{fryxell_multiple_2008}. 
\cite{schliehe-diecks_application_2012}
consider temporal trends in behavioural transitions over the
time scales of a six-hour
observation period, but
in general ecologists have turned to other tools to describe behavioural
changes over longer (diurnal, seasonal, or ontogenetic) time scales.
\todo{Fleming et al Am Nat 2014?  Gurarie?}

For movement behaviours that change on a fast time scale, such
that movement behaviours recorded
at successive observations are effectively independent,
\emph{finite mixture models} (FMM) --- which can be considered
a special case of HMMs where the
probability of state occupancy is independent of the previous state ---
can adequately describe movement \citep{tracey_mapping_2012}.  
When movement varies over long time scales
(relative to the time between observations) with little short-term
persistence or correlation, movement could be well represented by FMMs
where the occupancy probabilities change deterministically over time.
Thus FMMs and HMMs,
with or without temporal variation in the occupancy or transition
probabilities, form a useful family of models for capturing
changes in movement behaviour over a range of time scales.
\todo{do we want a figure showing this?}

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers.

@@

\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as 6 for individuals with a large amount
of available data); for practical reasons, they restricted their
detailed analysis to models with only 3 underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\citep{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\citep{dean2012behavioural,morales_extracting_2004,franke_prediction_2006,kerk2015hidden}. As \cite{kerk2015hidden} comment, behavioural
repertoires with more than three distinct states are difficult to
interpret, one possible reason that other authors have not adopted
\citeauthor{kerk2015hidden}'s model-based approach to
identifying the number of latent states.

@@

@@

Among the different types of time inhomogeneous transition HMMs, sinusoidal time dependent transition HMMs yields the lowest BIC. The multiple block transitions was a good fit for the male panthers, using a combination of three transition matrix ($b_{1}$ for $t={1-6,21-24}$, $b_{2}$ for $ t={7-16}$, and $b_{3}$ for $t={17-20}$) and it reduced the number of states by 1 but had no effect on the females. The quadratic and sinusoidal time covariate transitions was motivated by the multiple block transition pattern attempting to capture the cyclic diurnal cycle and same complex level in terms of number of parameters. We concluded that sinusoidal time dependent transition HMM was suitable for modeling these panthers.
